Containers and R---the Rockerverse and beyond

### 2020-01-15 manuscript.Rmd
add section on development and debugging
closes #41

Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not often needed.
Developers can readily start a local container based on a Rocker image to investigate a bug report, which might require them to install a specific combination of package versions or use a specific R version (see also versioned images above).
Unlike recreating the reporter's environment on their regular machine, a container can simply be discarded after use, and there is no need to tediously roll back explorative changes to a previous state.
Using the Rocker images with RStudio, the disposable environments lack no development comfort (see also the usage in research compendia).

**[`r-debug`](https://github.com/wch/r-debug)** is a purpose built Docker image for debugging R memory problems.
\citet{eddelbuettel_debugging_2019} describes how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).

The R-hub Docker images (see Section&nbsp;\nameref{rhub}) can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems they are not familiar.

>>>>

Containers can also serve as useful playgrounds to create environments ad-hoc or to provide very specific environments that are not needed or not easily available in day-to-day development.
These environments may have specific versions of R, of R extension packages, and of system libraries used by R extension packages, and all of the above in a specific combination.

First, such containers can greatly facilitate **fixing bugs**, because developers can readily start and later discard a container to investigate a bug report without affecting their regular system, and potentially having to revert changes back.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
For the [R-spatial community](https://www.r-spatial.org/), using containers is particularly useful, because of the strong integration of system libraries in core packages for geospatial data modelling and analysis.
Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries \footnote{\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, 
\href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial_proj6}, \href{https://github.com/r-spatial/sf/issues/1231#issuecomment-570076440}{https://github.com/r-spatial/sf/issues/1231\#issuecomment-570076440}}.

There are also special images for identifying problems beyond the mere R code, such as **debugging R memory problems**.
The images significantly reduce the barrier to follow complex steps for fixing memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
These problems are hard to debug and critical, both because when they occur they lead to fatal crashing processes.
[`rocker/r-devel-san`](https://github.com/rocker-org/r-devel-san) and [`rocker/r-devel-ubsan-clang`](https://github.com/rocker-org/r-devel-san-clang) are Docker images have a particularly configured version of R to trace such problems with gcc and clang compilers, respectively \citep[cf.~\pkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
The image [`wch/r-debug`](https://github.com/wch/r-debug) is a purpose built Docker image with _multiple_ instrumented builds of R, each with a different diagnostic utility activated.

Second, containers are useful for **testing** R code during development.
To submit a package to CRAN, an R package must work with the development version of R, which must be compiled locally.
That can be a challenge for some users.
The R-hub service (see Section&nbsp;\nameref{rhub}) makes it easy to ensure that no errors occur, but to fix errors a local setup is still often warranted, e.g. using the image `rocker/r-devel`, and to test packages with native code, which can make the process more complex \citep[cf.][]{eckert_building_2018}.
The R-hub Docker images can also be used to debug problems locally using various combinations of Linux platforms, R versions, and compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or _flavours_, used by CRAN for checking packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}}, e.g., with CentOS-based images, but lack a container for checking on Windows or OS X.
The images greatly support package developers to provide support on operating systems they are not familiar.
The package \pkg{dockertest} ([https://github.com/traitecoevo/dockertest/](https://github.com/traitecoevo/dockertest/)) is a proof of concept for automatically generating `Dockerfile`s and building images specifically to run tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied with a special launch script so the tested source code is not stored in the image but the currently checked in version from a local Git repository is cloned into the container at runtime.
This approach clearly seperates test environment, test code, and current working copy of the code.

Third, Docker images can be used **on CI platforms** to streamline the testing of packages.
\citet{ye_docker_2019} describes how they speed up the process of testing by running tasks on [Travis&nbsp;CI](https://travis-ci.org/) within a container using `docker exec`, e.g., the package check or rendering of documentation.
\citet{cardozo_faster_2018}, also on Travis&nbsp;CI saved time by re-using the testing image as the base iamge for an image intended for publication on Docker Hub.
Especially for long running tests or complex system dependencies, these approaches seems to improve the development.
Not due to a concern about time, but to control the environment used on a CI server, even this manuscript was rendered into a PDF and deployed to a GitHub-hosted website (see `.travis.yml` and `Dockerfile` in the manuscript repository).
This gives on the one hand easy access after every update of the R Markdown source code, and on the other hand a second controlled environment making sure that the article renders successfully and correctly.
